The evolution of gene expression in the term placenta of viviparous mammals

### 2015-12-18 manuscript/manuscript.Rnw
use macros for species names

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \textit{Homo sapiens} are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in humans is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven pregnancy-specific beta-1-glycoprotein genes are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of humans, with \textit{PSG1} showing the
greatest degree of expression %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

\subsection{Bonobo Placenta Transcriptome}

@@

\subsection{Spider Monkey Transcriptome}

\begin{table}
  \tiny
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.spider.monkey.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Ateles fusciceps}}
\label{tab:top50_spider_monkey_placenta}
\end{table}

@@

\begin{table}
  \tiny
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.dog.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Canis familiaris}}
\label{tab:top50_dog_placenta}
\end{table}

The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
in the term placenta of \dog are shown in
\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
expression %
(\ac{FPKM}=%
\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
among the top 50 (\cref{tab:top50_dog_placenta}) include genes
involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
inhibition of matrix metalloproteinases (\textit{TIMP2},
\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
immunosupression during pregnancy
\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
In addition, an ortholog of SSLP1 (ENSCAFG00000010638) a gene
expressed in the placentas of \cow (%
\Sexpr{gene.expression("bos tarus","SSLP1")}, %
\cref{tab:top50_cow_placenta}) and \sheep
(\Sexpr{gene.expression("ovis aries","SSLP1")},
\cref{tab:top50_sheep_placenta}) but is not present in primates
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
%
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
There is no placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in \dog (\cref{fig:estrogen_evolution_tree}), 
however expression of \textit{HSD3B1} is relatively high
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Mus musculus} are shown in
\cref{tab:top50_mouse_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_mouse_placenta}.

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Spalax carmeli} are shown in
\cref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_s_carmeli_placenta}.

@@

\subsection{Horse Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
top.horse.genes <- top.genes.table(species.name="equus caballus")
print(xtable(top.horse.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.horse.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Equus caballus}}
\label{tab:top50_horse_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Equus Caballus} are shown in
\cref{tab:top50_horse_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_horse_placenta}.

\subsection{Sheep Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
top.sheep.genes <- top.genes.table(species.name="ovis aries")
print(xtable(top.sheep.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.sheep.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Ovis aries}}
\label{tab:top50_sheep_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Ovis aries} are shown in
\cref{tab:top50_sheep_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_sheep_placenta}.

>>>>

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in humans is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven pregnancy-specific beta-1-glycoprotein genes are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

\subsubsection{\bonobo Placenta Transcriptome}

@@

\subsubsection{\smonkey Transcriptome}

\begin{table}
  \tiny
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.spider.monkey.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes in \smonkey which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene}
\label{tab:top50_spider_monkey_placenta}
\end{table}

@@

\begin{table}
  \tiny
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.dog.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of the top 50 genes in \dog which are
  not ribosomal, mitochondrial, or whose human ortholog is a
  housekeeping gene}
\label{tab:top50_dog_placenta}
\end{table}

The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
in the term placenta of \dog are shown in
\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
expression %
(\ac{FPKM}=%
\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
is 
(\ac{FPKM}=%
\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
among the top 50 (\cref{tab:top50_dog_placenta}) include genes
involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
inhibition of matrix metalloproteinases (\textit{TIMP2},
\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
immunosupression during pregnancy
\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
In addition, an ortholog of SSLP1 (ENSCAFG00000010638) a gene
expressed in the placentas of \cow (%
\Sexpr{gene.expression("bos tarus","SSLP1")}, %
\cref{tab:top50_cow_placenta}) and \sheep
(\Sexpr{gene.expression("ovis aries","SSLP1")},
\cref{tab:top50_sheep_placenta}) but is not present in primates
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
%
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
There is no placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in \dog (\cref{fig:estrogen_evolution_tree}), 
however expression of \textit{HSD3B1} is relatively high
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
\cref{tab:top50_mouse_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_mus_musculus_placenta}.

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Spalax carmeli} are shown in
\cref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_spalax_carmeli_placenta}.

@@

\subsection{\horse Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
top.horse.genes <- top.genes.table(species.name="equus caballus")
print(xtable(top.horse.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.horse.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes in \horse which are
  not ribosomal, mitochondrial, or whose human ortholog is a
  housekeeping gene}
\label{tab:top50_horse_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Equus Caballus} are shown in
\cref{tab:top50_horse_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_equus_caballus_placenta}.

\subsection{\sheep Placenta Transcriptome}

\begin{table}[h]
  \tiny
<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
top.sheep.genes <- top.genes.table(species.name="ovis aries")
print(xtable(top.sheep.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.sheep.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes in \sheep which are
  not ribosomal, mitochondrial, or whose human ortholog is a
  housekeeping gene}
\label{tab:top50_sheep_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \sheep are shown in
\cref{tab:top50_sheep_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_ovis_aries_placenta}.
